---
title: "R Notebook"
output: html_notebook
---

# Meta analysis of case-control studies

Case-control studies have binary outcomes and present event rate data. Usually, different studies analyze different populations, therefore a random-effects-model is appropriate. Also, in a case-control study, a Mantel-Haenszel method is appropriate to pool results, when you have raw data.

```{r}
library(readxl)
cannabis <- read_xlsx("Modelo.xlsx")
str(cannabis)
```

The raw data here is presented in:

Ee (events in experimental treatment arm)
Ne (N in experimental treatment arm)
Ec (events in control arm)
Nc (N in control arm)

And the pre-calculated data is presented in:

OR (odds ratio)
lower (lower bounds of the confidence interval)
upper (upper bounds of the confidence interval)

## Raw data

We can use the *meta* package to calculate the meta-analysis and present the **pooled Odds Ratio** like this:

```{r}
library(meta)
m.bin <- metabin(Ee,
                 Ne,
                 Ec,
                 Nc,
                 data = cannabis,
                 studlab = paste(Author),
                 comb.fixed = FALSE,
                 comb.random = TRUE,
                 method.tau = "SJ",
                 hakn = TRUE,
                 prediction = TRUE,
                 incr = 0.1,
                 sm = "OR")
m.bin
```

Now we will plot the meta-analysis in two ways.

```{r}
labbe.metabin(x = m.bin,
              bg = "blue",
              studlab = TRUE,
              col.random = "red")

# Forest plot
forest(m.bin, smlab = "Odds of lung cancer in\never cannabis smokers")
```

## Pre calculated data

First select only eligible studies that adjusted the effect size for smoking status, then calculate the log of the estimates.
```{r}
pre.cannabis <- cannabis[which(cannabis$Eligible_cigarette=="Yes"),]
pre.cannabis$OR <- log(pre.cannabis$OR)
pre.cannabis$lower <- log(pre.cannabis$lower)
pre.cannabis$upper <- log(pre.cannabis$upper)
```

Then, the standard error can be calculated and used to complete the meta-analysis.
```{r}
pre.cannabis$seTE <- (pre.cannabis$upper - pre.cannabis$lower)/3.92

m.gen <- metagen(OR,
                 seTE,
                 studlab = Author,
                 method.tau = "SJ",
                 sm = "OR",
                 data = pre.cannabis,
                 comb.fixed = F)
m.gen
```

And the plots can be shown as before:

```{r}
# Forest plot
forest(m.gen, smlab = "OR adjusted for\ncigarette smoking")
```

